Implementing simplified INLA into `aghq`

### 2022-11-11 src/dev_sinla/sinla.Rmd
Working on Laplace plus integration strategy

The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
$$
\begin{equation}
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), (\#eq:pxgiveny)
\end{equation}
$$
where $\theta(z) \in \Theta$ and $w_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
be the set of nodes in one-dimension, exactly the zeros of the Hermite polynomials $H_k(z_d)$ for $k \in \mathbb{N}$, and
$$
w_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
be the associated weights.
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ to be the set of nodes $z$ in $d$-dimensions, of which there will be $k^d$.
Then these points are adapted via $\theta(z) = \hat \theta + Lz$, where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.

This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue (in particular you solve a linear equation using the lower Cholesky triangle of the curvature at a randomly chosen quadrature point).
For the $i$th marginal, we just take the appropriate samples from the joint.
The approximation is therefore
$$
\begin{equation}
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z), (\#eq:pxigiveny)
\end{equation}
$$
where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mean vector and $(i, i)$th element of the covariance matrix.

@@

The goal of this notebook ([Implementing simplified INLA into `aghq`](https://athowes.github.io/elgm-inf/sinla.html)) is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
$$
\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
$$
We want to do this because the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
At the extreme, you could just compute a Laplace approximation for each marginal, integrating out all other elements of the latent field, $x_{-i}$, together with the fixed effects $\theta$.
This would be an accurate approach, but not computationally feasible for large models.
What we hope to achieve is an approximation which gets to most of the accuracy benefits of the Laplace approximation approach, but without the high computational cost.

@@

## Marginal of joint Gaussian approximation

@@

Now we want to evaluate
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) w_k(z).
$$

@@

The weight $w_k(z)$ is:

@@

To obtain the unlogged version we can use:
$$
\exp \left( \log\tilde p_\text{G}(x_i \, | \, \theta(z), y) + \log \tilde p_\text{LA}(\theta(z) \, | \, y) + \log w_k(z) \right)
$$

@@

Great -- now let's put it together and sum over `z` so that we can obtain the probability for our `x_test_value` in marginal `i`.
We won't worry about inefficient R for loops for the time being.

@@

```{r}
#' To-do
```

# Simplified INLA

In Section \@ref(full-laplace) we directly calculate the Laplace approximation to $\int p(x, y, \theta) \text{d} \theta_{-i}$.
Suppose that
$$
x \sim \mathcal{N}(\mu(\theta), Q(\theta)^{-1})
$$
then by the conditional distributions of a multivariate Gaussian distribution $x_{-i} \, | \, x_i$ is Gaussian with mean
$$
\mu(\theta)_{-i} + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \mu (\theta)_i) (\#eq:conditionalmean)
$$
and precision matrix
$$
Q(\theta)_{-i, -i} (\#eq:conditionalprecision)
$$

We follow Wood (2019) Section 2.
In this section we omit dependence on $\theta$ such that: $\mu = \mu(\theta)$, $Q = Q(\theta)$, $\mu(x_i) = \mu(x_i, \theta)$ and $Q(x_i) = Q(x_i, \theta)$.

Let $\mathcal{D}_{2j}$ denote a $2j \times 2j$ diagonal matrix with leading diagonal $-1, 1, -1, 1, \ldots$.
For example, with $j = 3$ we have:

```{r}
D2j <- function(j) {
  diag((-1)^{1:2 * j}, nrow = 2 * j)
}

D2j(3)
```

Let $u_0$ be a zero column matrix.
I am guessing that this means $u_j$ is a column matrix containing $j$ as each entry.
It's not specified what dimension $u_0$ or $u_j$ should be, but since the algorithm contains producing by $\mathcal{D}_{2j}$ presumably their lengths must be $2j$.

```{r}
uj <- function(j) {
 matrix(j, nrow = 2 * j, ncol = 1)
}

uj(3)
```

The algorithm contains $u_j \mathcal{D}_{2j} u_j^\top$.
If $u_j$ is a column matrix of dimension $\text{length}(u_j) \times 1$ then it's not going to be conformable with a $2j \times 2j$ matrix.
Is it actually a row matrix of dimension $1 \times \text{length}(u_j)$?
And $\text{length}(u_j) = 2j$ as discussed above?

Let $\Delta_j^i$ denote $\Delta_j$ with an extra zero inserted at element $i$.
But it hasn't been defined what $\Delta_j$ is yet.

## Step 1: Compute the Cholesky factor

The first task is to "compute the Cholesky factor of $Q_0 := Q_{-i, -i}$ by update of the Cholesky factor of $Q$".

```{r}
i <- 1

mm <- h$env$last.par
mean <- mm[h$env$random]
Q <- h$env$spHess(mm, random = TRUE)
Q0 <- Q[-i, -i]

L0 <- chol(Q0)
```

We have that $Q_0 = L_0 {L_0}^\top$:

```{r}
sum(abs((L0 %*% t(L0)) - Q0)) < 10E-9
```

## Step 2. Loop

For each $x_i$ in a grid of evaluation points (how does one chose this grid?) repeat the following:

```{r}
x_grid
```

### a)

Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean).
We start by computing the conditional mean for the first grid point:

```{r}
Sigma <- solve(Q)
conditional_mean <- mean[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mean[i])
```

Note that for this model the latent field is IID, so the conditional mean is the same as the mean.
Now we use Newton's method to find $\mu(x_i)$.
This requires some function $f(\mu(x_i), \ldots) = 0$.

### b)

Compute a set of $J$ steps $\{\Delta_j\}$

```{r}
J <- 1
```

### c)

```{r}
j <- 1
```

For $j = 1, \ldots, J - 1$ compute
$$
q = Q_0 \Delta_j + u_j \mathcal{D}_{2j}u_j^\top \Delta_j
$$
and
$$
g = \Delta_x \log p(\mu(x_i) + \Delta_j^i, y, \theta).
$$
Then compute the matrix
$$
u_{j + 1} = \{q(\Delta_j^\top j)^{-1/2}, g_{-i}(\Delta_j^\top g_{-i})^{-1/2}, u_j\}.
$$

### d)

Compute the determinant approximation
$$
\det(Q_1) = \det(Q_0) \det(I_{2J} + u_J^\top Q_0^{-1} u_J \mathcal{D}_{2j}).
$$

### e)

Compute $\tilde p(x_i \, | \, \theta, y) = \ldots$.

## Step 3. Renormalise $\tilde p(x_i \, | \, \theta, y)$

Renormalise $\tilde p(x_i \, | \, \theta, y)$.

>>>>

The AGHQ/ELGM approximation to the joint posterior of the latent field is the following object
$$
\begin{equation}
\tilde p(x \, | \, y) = \sum_z \tilde p_\text{G}(x \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z), (\#eq:pxgiveny)
\end{equation}
$$
where $\theta(z) \in \Theta$ and $\omega_k(z) \in \mathbb{R}$ are AGHQ nodes and weights generated by "adaption" to $\tilde p_\text{LA}(\theta \, | \, y)$.
In particular, let
$$
\mathcal{Q}(1, k) = \{z_d \in \mathbb{R}: H_k(z_d) = 0\}
$$
be the set of nodes in one-dimension, exactly the zeros of the Hermite polynomials $H_k(z_d)$ for $k \in \mathbb{N}$, and
$$
\omega_k(z_d) = \frac{k!}{H_{k + 1}(z_d)^2 \times \phi(z_d)}, 
$$
be the associated weights.
Define $\mathcal{Q}(d, k) = \mathcal{Q}(1, k)^d$ to be the set of nodes $z$ in $d$-dimensions, of which there will be $k^d$.
Then these points are adapted via $\theta(z) = \hat \theta + Lz$, where $\hat \theta$ is the mode (of $p_\text{LA}(\theta \, | \, y)$) and $L$ is the lower Cholesky triangle of the curvature (at the mode) such that $LL^\top = H^{-1}$.

This can be evaluated at any point, though in practice we work with it by drawing samples $x \sim p_\text{G}(x \, | \, \theta, y)$ using a method from sampling from a Gaussian given its mean and precision from Rue (in particular you solve a linear equation using the lower Cholesky triangle of the curvature at a randomly chosen quadrature point).
For the $i$th marginal, we just take the appropriate samples from the joint.
The approximation is therefore
$$
\begin{equation}
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z), (\#eq:pxigiveny)
\end{equation}
$$
where $p_\text{G}(x_i \, | \, \theta, y)$ just takes the $i$th elements of the mean vector and $(i, i)$th element of the covariance matrix.

@@

The goal of this notebook ([Implementing simplified INLA into `aghq`](https://athowes.github.io/elgm-inf/sinla.html)) is to create a new approximation swapping out $\tilde p_\text{G}(x_i \, | \, \theta, y)$ from the above for the simplified INLA approximation $p_{\text{SINLA}}(x_i \, | \, \theta, y)$ as follows
$$
\tilde p(x_i \, | \, \theta, y) = \sum_z \tilde p_{\text{SINLA}}(x_i \, | \, \theta, y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z).
$$
We want to do this because the new approximation should be more accurate than taking marginals of a joint Gaussian approximation.
At the extreme, you could just compute a Laplace approximation for each marginal, integrating out all other elements of the latent field, $x_{-i}$, together with the fixed effects $\theta$.
This would be an accurate approach, but not computationally feasible for large models.
What we hope to achieve is an approximation which gets to most of the accuracy benefits of the Laplace approximation approach, but without the high computational cost.

@@

## Marginal of joint Gaussian approximation {#aghq-gaussian-marginal}

@@

Now we want to evaluate
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{G}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z).
$$

@@

The weight $\omega_k(z)$ is:

@@

To obtain the unlogged version we can use:
$$
\exp \left( \log\tilde p_\text{G}(x_i \, | \, \theta(z), y) + \log \tilde p_\text{LA}(\theta(z) \, | \, y) + \log \omega_k(z) \right)
$$

@@

Great -- now let's sum over `z` so that we can obtain the probability for our `x_test_value` in marginal `i`.
We won't worry about inefficient R for loops for the time being.

@@

Let's do as we have above, but with the full Laplace approximation in place of the Gaussian.
We want to evaluate
$$
\tilde p(x_i \, | \, y) = \sum_z \tilde p_\text{LA}(x_i \, | \, \theta(z), y) \tilde p_\text{LA}(\theta(z) \, | \, y) \omega_k(z),
$$
so much of what we computed above in Section \@ref(aghq-gaussian-marginal) is still relevant.
The only thing that has changed is that we're now using $\tilde p_\text{LA}(x_i \, | \, \theta(z), y)$ rather than $\tilde p_\text{G}(x_i \, | \, \theta(z), y)$.

```{r}
x_test_value <- 0
i <- 1
z <- 1
```

As before, the weight $\omega_k(z)$ is:

And also as before, the normalised Laplace approximation $\tilde p_\text{LA}(\theta(z) \, | \, y)$ is:

Now, getting to the difference, the normalised Laplace approximation $p_\text{LA}(x_i \, | \, \theta(z), y)$ on the log-scale is:

```{r}
dat_i <- prepare_dat(dat, i)
map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))

for(theta in thetanames) param_i_fixed_theta[[theta]] <- modesandhessians[z, thetanames][[theta]]

#' Note that this is only a one-dimensional marginal, so perhaps we can afford to put k > 3 here
quad <- aghq::aghq(ff = obj_fixed_theta, k = 5, startingvalue = 0, control = aghq::default_control_tmb())

obj_fixed_theta_norm <- function(x) obj_fixed_theta$fn(x) + quad$normalized_posterior$lognormconst

obj_fixed_theta_norm(x_test_value)
```

Putting it together:

```{r}
#' On the log-scale
as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z])

#' Unlogged
exp(as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
```

Sum this over `z`:

for(z in 1:length(modesandhessians$weights)) {
  theta <- as.numeric(modesandhessians[z, thetanames])

  dat_i <- prepare_dat(dat, i)
  map_fixed_theta <- list(beta_prev = factor(NA), log_sigma_phi_prev = factor(NA))

  for(theta in thetanames) param_i_fixed_theta[[theta]] <- modesandhessians[z, thetanames][[theta]]

  #' Note that this is only a one-dimensional marginal, so perhaps we can afford to put k > 3 here
  quad <- aghq::aghq(ff = obj_fixed_theta, k = 5, startingvalue = 0, control = aghq::default_control_tmb())

  obj_fixed_theta_norm <- function(x) obj_fixed_theta$fn(x) + quad$normalized_posterior$lognormconst

  pdf <- pdf + exp(as.numeric(obj_fixed_theta_norm(x_test_value)) + (- h_norm(theta)) + log(modesandhessians$weights[z]))
}

<!-- # Simplified INLA -->

<!-- In Section \@ref(full-laplace) we directly calculate the Laplace approximation to $\int p(x, y, \theta) \text{d} \theta_{-i}$. -->
<!-- Suppose that -->
<!-- $$ -->
<!-- x \sim \mathcal{N}(\mu(\theta), Q(\theta)^{-1}) -->
<!-- $$ -->
<!-- then by the conditional distributions of a multivariate Gaussian distribution $x_{-i} \, | \, x_i$ is Gaussian with mean -->
<!-- $$ -->
<!-- \mu(\theta)_{-i} + \Sigma(\theta)_{-i, i} \Sigma(\theta)^{-1}_{i, i} (x_i - \mu (\theta)_i) (\#eq:conditionalmean) -->
<!-- $$ -->
<!-- and precision matrix -->
<!-- $$ -->
<!-- Q(\theta)_{-i, -i} (\#eq:conditionalprecision) -->
<!-- $$ -->

<!-- We follow Wood (2019) Section 2. -->
<!-- In this section we omit dependence on $\theta$ such that: $\mu = \mu(\theta)$, $Q = Q(\theta)$, $\mu(x_i) = \mu(x_i, \theta)$ and $Q(x_i) = Q(x_i, \theta)$. -->

<!-- Let $\mathcal{D}_{2j}$ denote a $2j \times 2j$ diagonal matrix with leading diagonal $-1, 1, -1, 1, \ldots$. -->
<!-- For example, with $j = 3$ we have: -->

<!-- ```{r} -->
<!-- D2j <- function(j) { -->
<!--   diag((-1)^{1:2 * j}, nrow = 2 * j) -->
<!-- } -->

<!-- D2j(3) -->
<!-- ``` -->

<!-- Let $u_0$ be a zero column matrix. -->
<!-- I am guessing that this means $u_j$ is a column matrix containing $j$ as each entry. -->
<!-- It's not specified what dimension $u_0$ or $u_j$ should be, but since the algorithm contains producing by $\mathcal{D}_{2j}$ presumably their lengths must be $2j$. -->

<!-- ```{r} -->
<!-- uj <- function(j) { -->
<!--  matrix(j, nrow = 2 * j, ncol = 1) -->
<!-- } -->

<!-- uj(3) -->
<!-- ``` -->

<!-- The algorithm contains $u_j \mathcal{D}_{2j} u_j^\top$. -->
<!-- If $u_j$ is a column matrix of dimension $\text{length}(u_j) \times 1$ then it's not going to be conformable with a $2j \times 2j$ matrix. -->
<!-- Is it actually a row matrix of dimension $1 \times \text{length}(u_j)$? -->
<!-- And $\text{length}(u_j) = 2j$ as discussed above? -->

<!-- Let $\Delta_j^i$ denote $\Delta_j$ with an extra zero inserted at element $i$. -->
<!-- But it hasn't been defined what $\Delta_j$ is yet. -->

<!-- ## Step 1: Compute the Cholesky factor -->

<!-- The first task is to "compute the Cholesky factor of $Q_0 := Q_{-i, -i}$ by update of the Cholesky factor of $Q$". -->

<!-- ```{r} -->
<!-- i <- 1 -->

<!-- mm <- h$env$last.par -->
<!-- mean <- mm[h$env$random] -->
<!-- Q <- h$env$spHess(mm, random = TRUE) -->
<!-- Q0 <- Q[-i, -i]  -->

<!-- L0 <- chol(Q0) -->
<!-- ``` -->

<!-- We have that $Q_0 = L_0 {L_0}^\top$: -->

<!-- ```{r} -->
<!-- sum(abs((L0 %*% t(L0)) - Q0)) < 10E-9 -->
<!-- ``` -->

<!-- ## Step 2. Loop -->

<!-- For each $x_i$ in a grid of evaluation points (how does one chose this grid?) repeat the following: -->

<!-- ```{r} -->
<!-- x_grid -->
<!-- ``` -->

<!-- ### a) -->

<!-- Use Newton's method with fixed Hessian $Q_{-i, -i}$ to find $\mu(x_i)$ starting from \@ref(eq:conditionalmean). -->
<!-- We start by computing the conditional mean for the first grid point: -->

<!-- ```{r} -->
<!-- Sigma <- solve(Q) -->
<!-- conditional_mean <- mean[-i] + Sigma[-i, i] * Sigma[i, i]^{-1} * (x_grid[1] - mean[i]) -->
<!-- ``` -->

<!-- Note that for this model the latent field is IID, so the conditional mean is the same as the mean. -->
<!-- Now we use Newton's method to find $\mu(x_i)$. -->
<!-- This requires some function $f(\mu(x_i), \ldots) = 0$. -->

<!-- ### b) -->

<!-- Compute a set of $J$ steps $\{\Delta_j\}$ -->

<!-- ```{r} -->
<!-- J <- 1 -->
<!-- ``` -->

<!-- ### c) -->

<!-- ```{r} -->
<!-- j <- 1 -->
<!-- ``` -->

<!-- For $j = 1, \ldots, J - 1$ compute -->
<!-- $$ -->
<!-- q = Q_0 \Delta_j + u_j \mathcal{D}_{2j}u_j^\top \Delta_j -->
<!-- $$ -->
<!-- and -->
<!-- $$ -->
<!-- g = \Delta_x \log p(\mu(x_i) + \Delta_j^i, y, \theta). -->
<!-- $$ -->
<!-- Then compute the matrix -->
<!-- $$ -->
<!-- u_{j + 1} = \{q(\Delta_j^\top j)^{-1/2}, g_{-i}(\Delta_j^\top g_{-i})^{-1/2}, u_j\}. -->
<!-- $$ -->

<!-- ### d) -->

<!-- Compute the determinant approximation -->
<!-- $$ -->
<!-- \det(Q_1) = \det(Q_0) \det(I_{2J} + u_J^\top Q_0^{-1} u_J \mathcal{D}_{2j}). -->
<!-- $$ -->

<!-- ### e) -->

<!-- Compute $\tilde p(x_i \, | \, \theta, y) = \ldots$. -->

<!-- ## Step 3. Renormalise $\tilde p(x_i \, | \, \theta, y)$ -->

<!-- Renormalise $\tilde p(x_i \, | \, \theta, y)$. -->
